---
title: ""
author: "Danielle Quinn"
date: "December 3, 2015"
output: html_document
---
This tutorial will briefly cover the basics of using Hadley Wickham's ggplot2 package, including generating basic plot layers and manipulating plotting aesthetics. Next, we'll go through a few complex but handy snippets of code that can improve your figure quality to publishing standards. This includes:

- setting consistent factor-specific aestethics
- facet wrap versus facet grid
- relabelling axes
- creating a shared legend
- aligning ggplot objects using grid arrange (get code from Trevor)
- plotting points behind boxplots
- melting data to facilitate factor-based aesthetics

This tutorial will use the ```mtcars``` data set:
```{r, comment=""}
data(mtcars)
```
You'll also need to load the `ggplot2` library:
```{r, comment=""}
library(ggplot2)
```
### Layers in ggplot2 ###
1. Set up your ggplot:
```{r, comment="", eval=FALSE}
ggplot(mtcars)
```
You'll notice that if you run this line of code, you'll get an error that says ```"No layers in plot"```. We'll fix this by adding a layer:
```{r, comment=""}
ggplot(mtcars)+
  geom_point(aes(x=cyl, y=wt))
```